### 2020-07-05 chapters/sota.tex
Multiple grammar corrections

\section{Outlier Detection in Time-Series Data} \label{sec:outliers}
Blázquez-García \emph{et al.} review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey, Aguinis-Outlier-survey, Chandola-Outlier-survey-2009, Hodge-Outlier-survey, Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.

In this work, the authors define a time-series as a collection of data points sorted by time; and outliers as observations that do not follow the expected behavior. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is “an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism" \cite{Hawkins-Outliers}. Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. For example, nearest neighbor based methods consider a datum an outlier if it has less than $k$ neighbors. Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

Quoting the authors, histogram-based univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.

@@

Yang et al. propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios. Data points are considered outliers if they have few neighbors, \textit{i.e.}, less than a user-defined threshold $\alpha$ of neighbors within a user-defined range $\theta$. Figure \ref{fig:point-outlier} depicts points plotted in their feature space (in this case only two dimensions or features are used, x1 and x2) and a point outlier for $alpha=2$ points and any given $\theta$.
\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=0.6]{figures/point-outlier.pdf}
    \caption{Point outlier for $\alpha=2$ and a given $\theta$}
    \label{fig:point-outlier}
  \end{center}
\end{figure}

@@

\subsubsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
Ahmad \& Purdy present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}. HTM has a few properties that are desirable for data stream anomaly detection scenarios. First, it is an unsupervised algorithm that requires no labels. This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent. Additionally, HTM adapts to noisy domains. In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored. Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.} changes in magnitudes of data or unusual timing for particular patterns, respectively.

@@

\subsubsection{DeepAnT}
Munir et. al present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}. DeepAnT uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series. This method is unsupervised - \textit{i.e.} it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

DeepAnT employs a Convolutional Neural Network (CNN) as its time-series predictor module. Before monitoring a data stream, the CNN model must be trained. The model can be trained with a small data set while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the data set used to train the model does not require labels.

\subsubsection*{Method classification}
DeepAnT takes as input a uni or multivariate time-series but is a univariate method by nature --- \textit{i.e.} it does not find correlations between multiple time-dependent variables. The model used by DeepAnT is a Convolutional Neural Network (CNN), used to make predictions on the future time-series values and measure differences with the actual observed values. Hence, DeepAnT is a prediction model-based univariate point outlier detection method (Figures \ref{fig:outlier-taxonomy} and \ref{fig:point-outlier-taxonomy}).

@@

with $F(S)$ as the frequency of subsequence $S$, $I(S)$ as the information carried by $S$ and $\alpha$ as a given threshold. According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}. The authors explain that if \textit{F(S)} is large --- \textit{i.e.} S occurs frequently --- then \textit{I(S)} will be closer to 0.

@@

\subsubsection*{Applicability to our Hypothesis}
In this Thesis, we want to detect and alert data pattern shifts with a low memory footprint. The method proposed by Kifer et. al grows linearly with window size, making it unfit for large windows, our use-case.

However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one. Since in this Thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.

@@

Sampaio et al. present SAMM \cite{SAMM}, an automatic model monitoring system for data streams. Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T. Window T contains the last $n_t$ elements and window R the $n_r$ elements of a reference period, which is prior to T. SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.

@@

As discussed in Section \ref{sec:back-swag-algs}, a sliding window aggregation algorithm allows the computation of an exact or approximate aggregation over a sliding window. We want to find an aggregation that is suitable to encode our reference and streaming periods so we avoid keeping the entire sliding window in memory. Furthermore, we need an efficient algorithm to incrementally maintain this aggregation. We have presented Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE) as the two most basic generic stream processing algorithms. In this Section we analyze other more sophisticated SWAG algorithms that fit different use-cases, restraining the set of computable aggregations to certain families, as defined in Table \ref{tbl:aggregations-properties}.

\subsection{Two-Stacks} \label{sec:2stacks}
The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.} elements are evicted in the same order they were inserted. Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B. Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack. The pseudocode for Two-Stacks can be seen in Algorithm \ref{pseudo:2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x. For example, if $\oplus$ is the arithmetic sum then $\overline{\theta}$ is 0. If $\oplus$ is the arithmetic multiplication then $\overline{\theta}$ is 1.

@@

\begin{algorithm}
    \caption{Two-Stacks insert, evict and query methods}
    \label{pseudo:2-stacks}
    \begin{algorithmic}[1]
        \Function{getStackAggregationValue}{stack}
         \If{stack.isEmpty()} 
            \State \Return{$\overline{\theta}$} 
        \Else 
            \State \Return{stack.top().agg} 
        \EndIf
        \EndFunction

@@

Inserting a new element corresponds to pushing a new value \textit{val} in \textit{vals} and pushing the new aggregation value \textit{agg} in \textit{aggs}. The new aggregation value \textit{agg} is computed by applying the \textit{combine} function $\oplus$ to the back element of \textit{aggs} and the new value \textit{val} --- \textit{i.e.} pushing \textit{val} to \textit{vals} and $\textit{aggs[E]} \oplus \textit{val}$ to \textit{aggs}.

@@

\textit{l\textsubscript{F}} and \textit{l\textsubscript{B}} combined cover the whole \textit{vals} queue --- \textit{i.e.} every element in the sliding window. \textit{l\textsubscript{F}} is aggregated to the left, thus \textit{aggs[F]} contains the aggregation value of everything in \textit{l\textsubscript{F}}. On the other hand, \textit{l\textsubscript{B}} is aggregated to the right, thus \textit{aggs[E]} stores the aggregation value of everything in \textit{l\textsubscript{B}}. Hence, the aggregation value of the entire window will be the result of \textit{$aggs[F] \oplus aggs[E]$}.

@@

Algorithm \ref{pseudo:daba} shows the pseudocode presented by the authors for the DABA algorithm. In reality, after each insertion and eviction, besides the procedure already described, function \textit{fixup} is executed. The \textit{fixup} function performs the incremental reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}}. Algorithms \ref{pseudo:daba-helpers} and \ref{pseudo:daba} show how such reversal is done but we will not elaborate on such procedure. However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across run time, in the form of DABA's \textit{fixup}.

\begin{algorithm}
    \caption{DABA helper functions}
    \label{pseudo:daba-helpers}
    \begin{algorithmic}[1]
        \Function{getListFAggregationValue}{}
            \If{F == B}
                \State \Return{$\overline{\theta}$}
            \Else
                \State \Return{aggs[F]}
            \EndIf
        \EndFunction

@@

\begin{algorithm}
    \caption{DABA insert, evict and query methods}
    \label{pseudo:daba}
    \begin{algorithmic}[1]
        \Function{query}{}
            \State \Return{getListFAggregationValue() $\oplus$ getListBAggregationValue()}
        \EndFunction

@@

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state-of-the-art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection{Membership Queries and the Bloom Filter} \label{sec:bloom}
% approx aggregation to compute
A Bloom Filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.} saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.} saying \textit{el} is not in the set but it actually is. In other words, a Bloom filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

% data structures
\subsection*{Data structure}
A Bloom filter is an array of bits of size \textit{m}, initially all set to 0. Figure \ref{fig:initial-bloom-filter} represents a Bloom filter with a bit array of size \textit{m}=7 in its initial state.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.9]{figures/initial-bloom.png}
      \caption[Bloom filter initial state]{Bloom filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

% algorithm for insertion
\subsection*{Element Insertion}
Element insertion and query answering make use of \textit{k} hash functions, each mapping elements to a position in the bit array. Adding an element is done by determining which positions should be set to 1. To that end, the new element is fed into each hash function that maps the element to an array position. The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1. Figure \ref{fig:insertion-bloom-filter} shows the insertion of element \textit{el} in the Bloom filter, making use of three hash functions (\textit{k}=3), \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}}. The outputs of the hash functions correspond to the array indexes to set to 1.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/insert-bloom.png}
      \caption[Bloom filter insertion]{Insertion of element \textit{el} into a Bloom filter with \textit{m}=7 and \textit{k}=3}
      \label{fig:insertion-bloom-filter}
    \end{center}
\end{figure}

% algorithm for query
\subsection*{Membership Query}
Performing a membership query --- \textit{i.e.} testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions. If \textit{el} was already in the set, all \textit{k} positions should be 1. If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/query-bloom.png}
      \caption[Bloom filter membership query]{Bloom filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

If all positions are 1 then the element is said to be \textit{possibly in the set}. The reason a Bloom filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements. For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$. This way, bits in positions $[1,2,3,4,5,6]$ are set to 1. When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp}. Since all of these positions are 1 because of the previous elements inserted, the Bloom filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set. This is considered a \textit{false positive}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
      \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
      \caption{Bloom filter false positive}
      \label{fig:bloom-filter-fp}
    \end{center}
\end{figure}

@@

Mullin presents an analysis of the Bloom filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
\begin{equation}
    \frac{1}{m}
\end{equation}
because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
\begin{equation}
    1-\frac{1}{m}
\end{equation}

@@

\subsection*{Optimal number of hash functions}
We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate. In order to do so, we must first allocate \textit{m} bits for the Bloom filter, choosing \textit{m} based on available memory. The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by:
\begin{equation}
    \frac{m}{n}ln(2)
\end{equation}

@@

Note that the time complexity does not at all depend on the number of elements processed by the Bloom filter and that \textit{k} will be a rather small constant. Hence, the Bloom filter is said to have constant time complexity.

\subsection*{Space complexity analysis}
Considering a Bloom filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity. Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit. Thus the Bloom filter has a constant space complexity.

\subsection*{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making them possible methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}. However, we need to evict old events, and for that reason, bloom filters are of no use to us.

\subsection{Item Frequency and the Count-Min Sketch}
% approx aggregation to compute
The Count-Min Sketch (CMS) uses two principles from the discussed Bloom filters in Section \ref{sec:bloom}. First, Bloom filters show that precision can be sacrificed to achieve space savings. Second, they are very similar at a technical level.

A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the data set. It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).

@@

\subsection*{Applicability to our Hypothesis}
Count-Min Sketches are constant in both time and space. Similarly to Bloom filters, this makes them ideal candidates for our solution.

@@

\subsection*{Space complexity analysis}
Quoting Flajolet et al. about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}

>>>>

\section{Outlier Detection in Time-Series Data} \label{sec:outliers}
Blázquez-García \emph{et al.} review state of the art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey, Aguinis-Outlier-survey, Chandola-Outlier-survey-2009, Hodge-Outlier-survey, Xu-outlier-survey-2019} we chose to adopt the taxonomy presented by Blázquez-García \emph{et al.} due to its focus on time-series data.

In this work, the authors define a time-series as a collection of data points sorted by time; and outliers as observations that do not follow the expected behavior. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is “an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism" \cite{Hawkins-Outliers}. Furthermore, Blázquez-García \emph{et al.} allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this thesis all refer to outliers. In summary, outlier detection refers to the process of finding anomalies in data. An anomaly (or outlier) relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

Density-based univariate point outlier detection methods rely on the clustering of data points and neighbor counting. For example, nearest neighbor based methods consider a datum an outlier if it has less than $k$ neighbors. Distance between two points is usually measured using the Euclidean distance \cite{elementary-algebra}. Two points are said to be neighbors if they are close to each other --- \textit{i.e.}, the distance between them is less than or equal to a given threshold $\alpha$.

The authors provide little detail on the last category, the histogram-based one. The authors state that these methods are ``based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original''.

@@

Yang \emph{et al.} propose \emph{Extra-N} \cite{Yang-Neighbor-Based-Pattern-Detection}, a method for incremental detection of point outliers in sliding window scenarios. Data points are considered outliers if they have few neighbors, \textit{i.e.}, less than a user-defined threshold $\alpha$ of neighbors within a user-defined range $\theta$. Figure \ref{fig:point-outlier} depicts points plotted in their feature space (in this case only two dimensions or features are used, x1 and x2) and a point outlier for $alpha=2$ points and any given $\theta$.
\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=0.6]{figures/point-outlier.pdf}
    \caption[Density-based point outlier]{Density-based point outlier for $\alpha=2$ and a given $\theta$}
    \label{fig:point-outlier}
  \end{center}
\end{figure}

@@

\subsubsection{Real-Time Stream Anomaly Detection with Hierarchical Temporal Memory}
Ahmad \& Purdy present an anomaly detection technique based on an online algorithm called Hierarchical Temporal Memory (HTM)\footnote{https://github.com/numenta/nupic} \cite{Ahmad-HTM}. HTM has a few properties that are desirable for data stream anomaly detection scenarios. First, it is an unsupervised algorithm that requires no labels. This property comes in handy when monitoring a data stream in real-time, where labels for incoming data will likely be absent. Additionally, HTM adapts to noisy domains. In other words, random variations in the underlying distribution of incoming data that do not persist throughout time will be ignored. Furthermore, HTM detects both spatial and temporal anomalies --- \textit{i.e.}, changes in magnitudes of data or unusual timing for particular patterns, respectively.

@@

\subsubsection{DeepAnT}
Munir \emph{et al.} present DeepAnT, a deep learning-based approach for the detection of anomalies in time-series data \cite{Munir-DeepAnT}. DeepAnT uses unlabeled data to capture and learn the distribution of the data stream used to forecast the normal behavior of the time-series. This method is unsupervised - \textit{i.e.}, it requires no labels - which is an advantage in a streaming scenario where real-time decisions have to be made before labels are effectively-known. It is capable of detecting point anomalies in time series data with periodic and seasonal characteristics.

DeepAnT employs a Convolutional Neural Network (CNN) as its time-series predictor module. Before monitoring a data stream, the CNN model must be trained. The model can be trained with a small dataset while achieving good generalization capabilities due to the effective parameter sharing of the CNN. Additionally, the dataset used to train the model does not require labels.

\subsubsection*{Method classification}
DeepAnT takes as input a uni or multivariate time-series but is a univariate method by nature --- \textit{i.e.}, it does not find correlations between multiple time-dependent variables. The model used by DeepAnT is a Convolutional Neural Network (CNN), used to make predictions on the future time-series values and measure differences with the actual observed values. Hence, DeepAnT is a prediction model-based univariate point outlier detection method (Figures \ref{fig:outlier-taxonomy} and \ref{fig:point-outlier-taxonomy}).

@@

with $F(S)$ as the frequency of subsequence $S$, $I(S)$ as the information carried by $S$ and $\alpha$ as a given threshold. According to the authors, \textit{"I(S) is computed taking into account the number of times the symbols within S are repeated through the series"}. The authors explain that if \textit{F(S)} is large --- \textit{i.e.}, S occurs frequently --- then \textit{I(S)} will be closer to 0.

@@

\subsubsection*{Applicability to our Hypothesis}
In this thesis, we want to detect and alert data pattern shifts with a low memory footprint. The method proposed by Kifer \emph{et al.} grows linearly with window size, making it unfit for large windows, our use case.

However, this two-window paradigm encodes a static normal or reference period in one window and the observed streaming values in a sliding one. Since in this thesis we want to report streaming data pattern shifts relatively to a reference period, we use a similar two-windowed model, but use lightweight and incrementally maintainable aggregations over both windows.

@@

Sampaio \emph{et al.} present SAMM \cite{SAMM}, an automatic model monitoring system for data streams. Similarly to the method presented in Section \ref{kifer}, SAMM uses two windows: the reference window R and the target window T. Window T contains the last $n_t$ elements and window R the $n_r$ elements of a reference period, which is prior to T. SAMM has two main components that interest us: the one responsible for computing the signal and the one that defines the signal threshold.

@@

As discussed in Section \ref{sec:back-swag-algs}, a sliding window aggregation algorithm allows the computation of an exact or approximate aggregation over a sliding window. We want to find an aggregation that is suitable to encode our reference and streaming periods so we avoid keeping the entire sliding window in memory. Furthermore, we need an efficient algorithm to incrementally maintain this aggregation. We have presented Recalculate-From-Scratch (RFS) and Subtract-On-Evict (SOE) as the two most basic generic stream processing algorithms. In this section, we analyze other more sophisticated SWAG algorithms that fit different use cases, restraining the set of computable aggregations to certain families, as defined in Table \ref{tbl:aggregations-properties}.

\subsection{Two-Stacks} \label{sec:2stacks}
The Two-Stacks SWAG algorithm works for \textit{associative} aggregations (recall Table \ref{tbl:aggregations-properties}) in scenarios where the sliding window has First-In First-Out (FIFO) semantics --- \textit{i.e.}, elements are evicted in the same order they were inserted. Two-Stacks \cite{Tangwongsan-DABA} implements the FIFO queue using two stacks, namely the front stack F and the back stack B. Each element in each of the stacks is a pair \textit{<val, agg>} where \textit{val} is the item's value and \textit{agg} is the aggregation value of everything below it on the stack. The pseudocode for Two-Stacks can be seen in Algorithm \ref{pseudo:2-stacks}, with $\oplus$ as the \textit{combine} function and $\overline{\theta}$ as the identity value of the \textit{combine} function, such that $\overline{\theta} \oplus x$ = $x \oplus \overline{\theta}$ = x. For example, if $\oplus$ is the arithmetic sum then $\overline{\theta}$ is 0. If $\oplus$ is the arithmetic multiplication then $\overline{\theta}$ is 1.

@@

\begin{algorithm}
    \caption[Two-Stacks]{Two-Stacks insert, evict and query methods}
    \label{pseudo:2-stacks}
    \begin{algorithmic}[1]
        \Function{getStackAggregationValue}{stack}
         \If{stack.isEmpty()} 
            \State \Return{$\overline{\theta}$} 
        \Else 
            \State \Return{stack.top().agg} 
        \EndIf
        \EndFunction

@@

Inserting a new element corresponds to pushing a new value \textit{val} in \textit{vals} and pushing the new aggregation value \textit{agg} in \textit{aggs}. The new aggregation value \textit{agg} is computed by applying the \textit{combine} function $\oplus$ to the back element of \textit{aggs} and the new value \textit{val} --- \textit{i.e.}, pushing \textit{val} to \textit{vals} and $\textit{aggs[E]} \oplus \textit{val}$ to \textit{aggs}.

@@

\textit{l\textsubscript{F}} and \textit{l\textsubscript{B}} combined cover the whole \textit{vals} queue --- \textit{i.e.}, every element in the sliding window. \textit{l\textsubscript{F}} is aggregated to the left, thus \textit{aggs[F]} contains the aggregation value of everything in \textit{l\textsubscript{F}}. On the other hand, \textit{l\textsubscript{B}} is aggregated to the right, thus \textit{aggs[E]} stores the aggregation value of everything in \textit{l\textsubscript{B}}. Hence, the aggregation value of the entire window will be the result of \textit{$aggs[F] \oplus aggs[E]$}.

@@

Algorithm \ref{pseudo:daba} shows the pseudocode presented by the authors for the DABA algorithm. In reality, after each insertion and eviction, besides the procedure already described, function \textit{fixup} is executed. The \textit{fixup} function performs the incremental reversal of \textit{l\textsubscript{B}} to \textit{l\textsubscript{F}}. Algorithms \ref{pseudo:daba-helpers} and \ref{pseudo:daba} show how such reversal is done but we will not elaborate on such procedure. However, it is important to stress out that this is the core idea of DABA: avoid occasional expensive operations, like Two-Stacks' \textit{flip}, by spreading them out across runtime, in the form of DABA's \textit{fixup}.

\begin{algorithm}
    \caption[De-Amortized Banker’s Aggregator (DABA) helpers]{DABA helper functions}
    \label{pseudo:daba-helpers}
    \begin{algorithmic}[1]
        \Function{getListFAggregationValue}{}
            \If{F == B}
                \State \Return{$\overline{\theta}$}
            \Else
                \State \Return{aggs[F]}
            \EndIf
        \EndFunction

@@

\begin{algorithm}
    \caption[De-Amortized Banker’s Aggregator (DABA)]{DABA insert, evict and query methods}
    \label{pseudo:daba}
    \begin{algorithmic}[1]
        \Function{query}{}
            \State \Return{getListFAggregationValue() $\oplus$ getListBAggregationValue()}
        \EndFunction

@@

\section{Probabilistic Data Structures} \label{sec:pds}
In Section \ref{sec:sota-swag-algs} we presented some state of the art algorithms for computing sliding window aggregations (SWAGs). However, if the aggregation we desire to compute can be approximated, there may exist more memory efficient methods. These aggregators are commonly denoted as sketches or probabilistic data structures and produce results orders-of-magnitude faster than some exact approaches while keeping the memory consumption constant and providing mathematically proven error bounds. Each probabilistic data structure is used for efficient computation of one approximate aggregation, hence we say they are aggregation-specific.

@@

\subsection{Membership Queries and the Bloom Filter} \label{sec:bloom}
% approx aggregation to compute
A Bloom Filter \cite{BLOOM-BLOOMFILTER} approximate aggregator allows membership queries on a set of elements. Bloom Filters answer the query: "Is element \textit{el} in the set of seen elements so far?". Being a probabilistic data structure and approximate aggregator, Bloom Filters' answers have an error associated. \textit{False positive} matches are possible --- \textit{i.e.}, saying \textit{el} is in the set when it is not --- but \textit{false negatives} are not --- \textit{i.e.}, saying \textit{el} is not in the set but it actually is. In other words, a Bloom Filter will accurately identify all items that do not belong in the set but will misclassify some items as being present in the set. Hence, a query returns either \textit{possibly in set} or \textit{definitely not in set}.

% data structures
\subsection*{Data structure}
A Bloom Filter is an array of bits of size \textit{m}, initially all set to 0. Figure \ref{fig:initial-bloom-filter} represents a Bloom Filter with a bit array of size \textit{m}=7 in its initial state.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.9]{figures/initial-bloom.png}
      \caption[Bloom Filter initial state]{Bloom Filter bit array initial state of \textit{m}=7}
      \label{fig:initial-bloom-filter}
    \end{center}
\end{figure}

% algorithm for insertion
\subsection*{Element Insertion}
Element insertion and query answering make use of \textit{k} hash functions, each mapping elements to a position in the bit array. Adding an element is done by determining which positions should be set to 1. To that end, the new element is fed into each hash function that maps the element to an array position. The resultant \textit{k} outputs are used as the \textit{k} positions of the array to be set to 1. Figure \ref{fig:insertion-bloom-filter} shows the insertion of element \textit{el} in the Bloom Filter, making use of three hash functions (\textit{k}=3), \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}}. The outputs of the hash functions correspond to the array indexes to set to 1.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/insert-bloom.png}
      \caption[Bloom Filter insertion]{Insertion of element \textit{el} into a Bloom Filter with \textit{m}=7 and \textit{k}=3}
      \label{fig:insertion-bloom-filter}
    \end{center}
\end{figure}

% algorithm for query
\subsection*{Membership Query}
Performing a membership query --- \textit{i.e.}, testing if an element \textit{el} is in the set --- is done by feeding \textit{el} into each one of the \textit{k} hash functions in order to get an array of positions. If \textit{el} was already in the set, all \textit{k} positions should be 1. If any position contains a 0 then the element is \textit{definitely not in set}, as shown in Figure \ref{fig:bloom-filter}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/query-bloom.png}
      \caption[Bloom Filter membership query]{Bloom Filter membership test of element \textit{el}. Finding at least one ``0" indicates that it is not in the set.}
      \label{fig:bloom-filter}
    \end{center}
\end{figure}

If all positions are 1 then the element is said to be \textit{possibly in the set}. The reason a Bloom Filter does not provide 100\% certainty that \textit{el} is in the set in case of all positions being 1 is that hash functions may give the same position for two different elements. For example, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map an element \textit{el1} to positions $[1,2,3]$ and an element \textit{el2} to positions $[4,5,6]$. This way, bits in positions $[1,2,3,4,5,6]$ are set to 1. When querying about the membership of a new element \textit{el3}, hash functions \textit{h\textsubscript{1}}, \textit{h\textsubscript{2}} and \textit{h\textsubscript{3}} may map it to positions $[1,3,5]$ as exemplified in Figure \ref{fig:bloom-filter-fp}. Since all of these positions are 1 because of the previous elements inserted, the Bloom Filter returns \textit{"possibly in set"} when in reality \textit{el3} was not in the set. This is considered a \textit{false positive}.

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/fp-bloom-1.png}
      \includegraphics[scale=0.4]{figures/fp-bloom-2.png}
      \includegraphics[scale=0.4]{figures/fp-bloom-3.png}
      \caption{Bloom Filter false positive}
      \label{fig:bloom-filter-fp}
    \end{center}
\end{figure}

@@

Mullin presents an analysis of the Bloom Filter in \cite{Mullin-Bloom-Analysis} where he details the false positive rate formula. The probability of one arbitrary bit being set to 1 is:
\begin{equation}
    \frac{1}{m}
\end{equation}
because there are \textit{m} bits. Likewise, the probability of one bit not being set to 1 is
\begin{equation}
    1-\frac{1}{m}
\end{equation}

@@

\subsection*{Optimal number of hash functions}
We want to choose the number of hash functions \textit{k} in order to minimize the false positive rate. In order to do so, we must first allocate \textit{m} bits for the Bloom Filter, choosing \textit{m} based on available memory. The value of \textit{k} that minimizes the false positive rate is either of the nearest integers given by:
\begin{equation}
    \frac{m}{n}ln(2)
\end{equation}

@@

Note that the time complexity does not at all depend on the number of elements processed by the Bloom Filter and that \textit{k} will be a rather small constant. Hence, the Bloom Filter is said to have constant time complexity.

\subsection*{Space complexity analysis}
Considering a Bloom Filter with \textit{m} bits, the space required is simply the array of \textit{m} bits, thus \textit{O(m)} space complexity. Similarly to the time complexity analysis, we point out that \textit{m} will be constant and that each of the array elements occupies 1 bit. Thus the Bloom Filter has a constant space complexity.

\subsection*{Applicability to our Hypothesis}
Bloom Filters exhibit constant space and time complexity making them possible methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}. However, we need to evict old events, and for that reason, Bloom Filters are of no use to us.

\subsection{Item Frequency and the Count-Min Sketch}
% approx aggregation to compute
The Count-Min Sketch (CMS) uses two principles from the discussed Bloom Filters in Section \ref{sec:bloom}. First, Bloom Filters show that precision can be sacrificed to achieve space savings. Second, they are very similar at a technical level.

A Count–Min Sketch \cite{Cormode-CMS} is an approximate aggregator that estimates the frequency of each element in the dataset. It is named after the two basic operations used to provide the estimate: counting (\textit{count}) and computing the minimum (\textit{min}).

@@

\subsection*{Applicability to our Hypothesis}
Count-Min Sketches are constant in both time and space. Similarly to Bloom Filters, this makes them ideal candidates for our solution.

@@

\subsection*{Space complexity analysis}
Quoting Flajolet \emph{et al.} about the memory consumption and error in the approximations obtained in their experiments in the original HyperLogLog paper \cite{Flajolet-HLL}: 
\say{cardinalities till values over N = $10^9$ can be estimated with a typical accuracy of 2\% using 1.5kB (kilobyte) of storage.}
